\bf On semidefinite representable sets over valuation fields

### 2024-11-11 article/paper_padic_spec.tex
commit by Simone

\newenvironment{proof}{\hbox{}\vspace{-0.8cm} {\bf Proof:}}{\hfill $\Box$}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathrsfs} 
\usepackage{graphicx}
\usepackage{color}
\usepackage{ulem}
\usepackage{bm}

@@

\textheight235mm
\textwidth165mm
\voffset-10mm
\hoffset-12.5mm
\parindent0cm
\parskip2mm

@@

%\title{\bf Title}
%\title{\bf On polyhedra and spectrahedra over valuation fields}
\title{\bf On semidefinite-representable sets over valuation fields}

@@

\begin{abstract}
  \noindent
  Real polyhedra and real spectrahedra are respectively the feasible sets of linear programming
  (LP) and semidefinite programming (SDP). This paper investigates versions of these sets and of
  the related optimization problems with data over a field $K$ with valuation.
  We showcase examples of non-polyhedral $K$-spectrahedra and of sets that are semidefinite
  representable over $K$ but are not $K$-spectrahedra.
\end{abstract}

@@

\subsection{Definition}
%\newcommand\mat{postive semidefinite matrix } 
\newcommand\Mat{Positive semidefinite matrix }
\newcommand\mats{positive semidefinite matrices }
\newcommand\Mats{positive semidefinite matrices }

@@

Recall that the {\it Newton polygon} $\mathcal{P}$ of a polynomial $P = \sum_{i=0}^{n} a_i T^i \in K[T]$ is
defined as the inferior convex hull of the set $\{(i, \val a_i) \,|\, i\in \{0,...,d\} \}$, {\it i.e.} the
graph of the greatest convex function $\mathcal{P} : [0,n] \to \R$ such that $\mathcal{P}(i) \leq \val a_i$
for $i=1,...,n$. This function is piecewise affine and its slopes are exactly the opposites of the valuations
of the roots of $P$.
  %%%These are all the results we will need for this proof.

Testing positive semidefiniteness reduces to checking the coefficients of $\chi_M$.
\begin{theorem}
  \label{caracsdp}
  $M \in \allmat_{d,d}^+(K)$ if and only if $\chi_M \in \OK[T]$.
\end{theorem}
\begin{proof}
  Both senses of the equivalence will be proven by contraposition.
  Let $M \in \allmat_{d,d}(K)$ with characteristic polynomial $\chi_M = T^d + \sum_{i=0}^{d-1} a_i T^i$ for $a_0,...,a_{d-1} \in K$.
  Firstly, suppose $M \not \in \allmat_{d,d}^+(K)$, {\it i.e.} $\chi_M$ has a root $\rho$ with $\val\,\rho<0$.
  It follows that
  $$
  \val \chi_M (\rho) = \val \left(\rho^d+ \sum_{i=0}^{d-1} a_i \rho^i\right) = + \infty
  $$ and $\val \rho^d = d \val \rho <0$.
  The properties of the non-archimedean inequality and of the valuation imply that
  $$
  \min_{0 \leq i \leq d-1} \val \left(a_i \rho^i\right) \leq \val\left(\sum_{i=0}^{d-1} a_i \rho^i\right) = \val \rho^d
  $$
  %%%$\val ( \sum_{i=0}^{d-1} a_i \rho^i) = \val \rho^d$ and $\val ( \sum_{i=0}^{d-1} a_i \rho^i) \ge \min_{i=0,..., d-1} \val \left(a_i \rho^i\right)$.
  Therefore, there exists $j \in \left\{0,...,d-1\right\}$ such that $\val(a_j \rho^j) \le \val\rho^d$,
  {\it i.e.} such that $\val a_j \le (d-j) \val \rho <0$. This shows that $\chi_M \not\in \OK[T]$.

@@

  Suppose now $\chi_M \not\in \OK[T]$. There exists $j \in \{0,...,d-1\}$ such that $\val a_j <0$. The polynomial $\chi_M$ being monic, the point with abscissa $n$ of the Newton polygon of $\chi_M$ is $(n,0)$. 
  Therefore, the point $(j, \val a_j)$ having negative ordinate implies that at least one of the slopes of
  the Newton polygon of $\chi_M$ is positive, so that $\chi_M$ has a root with negative valuation.
\end{proof}

@@

\begin{example}
  The reverse inclusion of \Cref{cor_caracsdp} is false in general. For instance the characteristic polynomial of
  $M = \left(\begin{smallmatrix} 5 + {3}/{5} & {4}/{5} \\ {4}/{5} & -{3}/{5} \end{smallmatrix}\right)$ is $\chi_M = T^2-5T-4$. However when seen as a matrix over $K=\mathbb{Q}_5$, then $\chi_M \in \mathbb{Z}_5[T] = \OK[T]$: then $M \in \allmat_{2,2}^+(\mathbb{Q}_5)$, nevertheless $M \not\in \allmat_{2,2}(\mathbb{Z}_5)$, hence $\allmat_{2,2}^+(\mathbb{Q}_5) \not\subset \allmat_{2,2}(\mathbb{Z}_5)$.
\end{example}

\begin{proposition}
  The set $\allmat_{d,d}^+(K)$ is both open and closed.
\end{proposition}
\begin{proof}
Firstly, $\OK[X]$ is closed and open in $K[X]$ equipped with the infinitiy norm $\|\cdot\|_\infty : P = \sum_{k=1}^{n} a_k X^k \to \sup |a_k|$. $\OK$ is indeed the open unit ball for $\|\cdot \|_\infty$ that induces a discrete distance for which open balls are also closed sets.
Furthermore, $\allmat^+_{d,d}(K) = \chi^{-1}(\OK[X]) $ where $\chi$ is the function that maps a matrix to its characteristic polynomial which is continuous.
\end{proof}

\begin{definition}
  A \emph{cone} is subset $C$ of $K^d$ that is closed under multiplication by elements of $\OK$.   
\end{definition}

@@

@@

\subsection{Polyhedrality}
Cf. \cite{bhardwaj2015deciding}

\section{Linear matrix inequalities}

\section{Other questions}
\begin{itemize}
\item The projection of a real polyhedron is a real polyhedron (thanks to Fourier-Motzkin elimination procedure). Is it the same for $K$-polyhedra?
\item Annuli are semidefinite representable, we should remark/prove that they are not spectrahedra.
\end{itemize}

>>>>

%%%\newenvironment{proof}{\hbox{}\vspace{-0.8cm} {\bf Proof:}}{\hfill $\Box$ \vspace{0.2cm}}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{mathrsfs} 
\usepackage{graphicx}
\usepackage{color}
\usepackage{ulem}
\usepackage{bm}

@@

\textheight235mm
\textwidth165mm
\voffset-10mm
\hoffset-12.5mm
\parindent0.8cm
\parskip0mm

@@

%\title{\bf Title}
%\title{\bf On polyhedra and spectrahedra over valuation fields}
\title{\bf On semidefinite representable sets over valuation fields}

@@

\begin{abstract}
  \noindent
  Real polyhedra and real spectrahedra (or their projections) are respectively the feasible sets
  of linear programming (LP) and semidefinite programming (SDP). 
  This paper investigates versions of these sets and of
  the related optimization problems with data over a field $K$ with valuation.
  We showcase examples of non-polyhedral $K$-spectrahedra and of sets that are semidefinite
  representable over $K$ but are not $K$-spectrahedra.
\end{abstract}

@@

\subsection{Positive semidefinite matrices}
%\newcommand\mat{postive semidefinite matrix } 
\newcommand\Mat{Positive semidefinite matrix }
\newcommand\mats{positive semidefinite matrices }
\newcommand\Mats{positive semidefinite matrices }

@@

Recall that the {\it Newton polygon} $\mathcal{P}$ of a polynomial $P = \sum_{i=0}^{d} a_i T^i \in K[T]$ is
defined as the inferior convex hull of the set $\{(i, \val a_i) \,|\, 0 \le i \le d\}$, {\it i.e.} the
graph of the greatest convex function $\mathcal{P} : [0,d] \to \R$ such that $\mathcal{P}(i) \leq \val a_i$
for $i=1,\ldots,d$. This function is piecewise affine and its slopes are exactly the opposites of the valuations
of the roots of $P$.

The following result shows that testing positive semidefiniteness of matrices over
valuation fields reduces to checking whether the coefficients of $\chi_M$ are integers.

\begin{theorem}
  \label{caracsdp}
  $M \in \allmat_{d,d}^+(K)$ if and only if $\chi_M \in \OK[T]$.
\end{theorem}
\begin{proof}
  Both senses of the equivalence will be proven by contraposition.
  Let $M \in \allmat_{d,d}(K)$ with characteristic polynomial $\chi_M = T^d + \sum_{i=0}^{d-1} a_i T^i$ for $a_0,...,a_{d-1} \in K$.
  First suppose $M \not \in \allmat_{d,d}^+(K)$, thus $\chi_M$ has a root $\rho$ with $\val\,\rho<0$.
  It follows that
  $$
  + \infty = \val \chi_M (\rho) = \val \left(\rho^d+ \sum_{i=0}^{d-1} a_i \rho^i\right)
  $$ and $\val \rho^d = d \val \rho <0$.
  The properties of the non-archimedean inequality and of the valuation imply that
  $$
  0> \val \rho^d = \val\left(\sum_{i=0}^{d-1} a_i \rho^i\right) \geq
  \min_{0 \leq i \leq d-1} \val \left(a_i \rho^i\right)
  $$
  %%%$\val ( \sum_{i=0}^{d-1} a_i \rho^i) = \val \rho^d$ and $\val ( \sum_{i=0}^{d-1} a_i \rho^i) \ge \min_{i=0,..., d-1} \val \left(a_i \rho^i\right)$.
  Therefore, there exists $j \in \left\{0,...,d-1\right\}$ such that $\val(a_j \rho^j) \le \val\rho^d$,
  {\it i.e.} such that $\val a_j \le (d-j) \val \rho <0$. This shows that $\chi_M \not\in \OK[T]$.

@@

  Suppose now $\chi_M \not\in \OK[T]$. There exists $j \in \{0,...,d-1\}$ such that $\val a_j <0$.
  The polynomial $\chi_M$ being monic, the point with abscissa $d$ of the Newton polygon of $\chi_M$ is
  $(d,\val(1)) = (d,0)$. 
  Therefore, the point $(j, \val a_j)$ having negative ordinate implies that at least one of the slopes of
  the Newton polygon of $\chi_M$ is positive, so that $\chi_M$ has a root with negative valuation.
\end{proof}

As a consequence, matrices with integer coefficients are positive semidefinite.

@@

The reverse inclusion of \Cref{cor_caracsdp} is false in general as we show in the next example.

\begin{example}
  Let
  $$
  M = \begin{pmatrix} 5 + \frac{3}{5} & \frac{4}{5} \\ \frac{4}{5} & -\frac{3}{5} \end{pmatrix}.
  $$
  The characteristic polynomial of $M$
  is $\chi_M = T^2-5T-4$. However when seen as a matrix over $K=\mathbb{Q}_5$, then $\chi_M \in \mathbb{Z}_5[T] = \OK[T]$: then $M \in \allmat_{2,2}^+(\mathbb{Q}_5)$, nevertheless $M \not\in \allmat_{2,2}(\mathbb{Z}_5)$, hence $\allmat_{2,2}^+(\mathbb{Q}_5) \not\subset \allmat_{2,2}(\mathbb{Z}_5)$.
\end{example}

\begin{proposition}
  {\color{blue} [Peut être une hypothèse à rajouter, par exemple $K$ à valuation discrete]}
  The set $\allmat_{d,d}^+(K)$ is both open and closed.
\end{proposition}
\begin{proof}
Firstly, $\OK[X]$ is closed and open in $K[X]$ equipped with the infinitiy norm $\|\cdot\|_\infty : P = \sum_{k=1}^{n} a_k X^k \to \sup |a_k|$. $\OK$ is indeed the open unit ball for $\|\cdot \|_\infty$ that induces a discrete distance for which open balls are also closed sets.
Furthermore, $\allmat^+_{d,d}(K) = \chi^{-1}(\OK[X]) $ where $\chi$ is the function that maps a matrix to its characteristic polynomial which is continuous.
\end{proof}

\begin{definition}
  A \emph{cone} of $K^d$ is a subset that is closed under multiplication by elements of $\OK$.   
\end{definition}

@@

\subsection{Definition of spectrahedra}

@@

%\subsection{Polyhedrality}
%Cf. \cite{bhardwaj2015deciding}

%\section{Linear matrix inequalities}

%\section{Other questions}
%\begin{itemize}
%\item The projection of a real polyhedron is a real polyhedron (thanks to Fourier-Motzkin elimination procedure). Is it the same for $K$-polyhedra?
%\item Annuli are semidefinite representable, we should remark/prove that they are not spectrahedra.
%\end{itemize}

\section*{Acknowledgments}

The second author acknowledges support from the ANR Project ``HYPERSPACE'' (ANR-21-CE48-0006-01).
